A thermodynamic view of dusty protoplanetary disks: axisymmetric stability

### 2016-10-13 model.tex
revising with youdin's comments

\section{Single fluid description of dusty gas}\label{setup} 
In the limit of strong dust-gas drag (stopping time $\tstop\to 0$) 
\cite{laibe14} give a single fluid description of the mixture:  
(their Eq. 84 --- 85 and Eq. 87):

\begin{align}
  &\frac{D\rho}{Dt} = -\rho\nabla\cdot\bm{v}, \label{masseq}\\ 
  &\frac{D\bm{v}}{Dt} = - \nabla\Phi - \frac{1}{\rho}\nabla  P, \label{momeq}\\ 
  &\frac{D\tepsilon}{Dt} = -\frac{1}{\rho} \nabla \cdot \left(\tepsilon 
  \tstop \nabla P \right),\label{dusteq}  
\end{align}
where $D/Dt \equiv \p_t + \bm{v}\cdot\nabla$ is the Lagragian
derivative; 
\begin{align}
  \rho \equiv \rhog + \rhod
\end{align}
is the total density with $\rhog$ and $\rhod$ being the gas and dust
density, respectively; 
\begin{align}
  \bm{v} \equiv \frac{\rhog\bm{v}_\mathrm{g} + 
    \rhod\bm{v}_\mathrm{d}}{\rho}
\end{align}
is the center-of-mass velocity with $\bm{v}_\mathrm{g}$ and
$\bm{v}_\mathrm{d}$ being the gas and dust velocities, respectively; 
and 
\begin{align}
  \tepsilon \equiv \frac{\rhod}{\rho}  = \frac{\epsilon}{1+\epsilon} 
\end{align}
is the dust-fraction and $\epsilon=\rhod/\rhog$ is the usual
dust-to-gas ratio. The gas pressure $P$ is given by the chosen 
equation of state described below. Note that Eq. \ref{dusteq} is
equivalent to Eq. 46 of \cite{jacquet11}.

The external potential $\Phi$ is 
that for a star of mass $M_*$, 
\begin{align}\label{thin_disk_potential}
  \Phi(r,z) =-\frac{GM_*}{\sqrt{r^2 + z^2}}\simeq
  -\frac{GM_*}{r}\left(1 - \frac{z^2}{2r^2}\right), 
\end{align}
where $G$ is the gravitational constant, and $(r,\phi, z)$ are
cylindrical co-ordinates centered on the star. 
The second equality is the 
thin-disk approximation for the disk potential, appropriate for
$|z|\ll r$. We shall adopt this approximate potential in order to
obtain analytic expressions for the disk equilibria.

\subsection{Locally isothermal equation of state}\label{loc_iso_eos}
We adopt 
\begin{align}\label{eos}
  P = c_s^2(r,z)\rhog = c_s^{2}(1 - \tepsilon)\rho,   
\end{align}
where $c_s$ is a prescribed sound-speed profile fixed in time. 
In practice we consider vertically 
isothermal disks with \begin{align}\label{power_temp}
  c_s^2(r) \propto r^{q},
\end{align}
where $q$ is the power-law index for the disk temperature. For $q=0$
the disk is strictly isothermal.

@@

\subsection{Effective energy equation}
We show that for a prescribed temperature distribution, the mixture
obeys an evolutionary energy equation, due to the advection of the
dust-fraction. Eliminate $\tepsilon$ in favor of $P$, 
\begin{align*}
  \tepsilon = 1 - \frac{P}{c_s^2(r,z)\rho}, 
\end{align*}
then Eq. \ref{dusteq} becomes
\begin{align}
\frac{DP}{Dt} &= - P \nabla\cdot\bm{v} + P\bm{v}\cdot\nabla\ln{c_s^2}
                + \mathcal{C},  \label{eff_energy} \\
\mathcal{C}&\equiv c_s^2 \nabla\cdot\left[\tstop\left(1 -
  \frac{P}{c_s^2\rho}\right)\nabla 
  P\right].
\end{align}
Eq. \ref{eff_energy} is the energy equation for an ideal gas of adiabatic index
$\gamma=1$, i.e. isothermal evolution, but now with additional energy source 
terms due to the imposed temperature profile via $\nabla c_s^2$, as
well as dust-gas drag via $\mathcal{C}$. 
%In this form, the function $\mathcal{C}$ can be interpreted
%as a cooling term.    
If we denote 
\begin{align}
  \bm{F} \equiv  - \frac{\nabla P}{\rho}
\end{align}
for the pressure forces, then
\begin{align*}
  \mathcal{C} = - c_s^2\nabla \left( \tepsilon \tstop \rho \bm{F}
  \right), 
\end{align*}
which is in the same form as cooling by radiative diffusion. In protoplanetary
disks the heat flux, proportional to $\bm{F}$, is directly radially
outwards and vertically upwards. This is simply a reflection of
particle drift (inwards and downwards). Particle flux into a region contributes
to `cooling' of that region because the effective temperature is
lowered.

>>>>

\section{Single fluid description of dusty gas}\label{setup} 
We model an accretion disk as a mixture of gas with dust treated as a 
pressureless fluid. We denote their density and
velocity field as $(\rhog,\bm{v}_\mathrm{g})$ and
$(\rhod,\bm{v}_\mathrm{d})$, respectively. The two fluids 
interact via a drag force parameterized by a stopping time $\tstop$
such that 
\begin{align}  
  \rhod\left.\frac{\p \bm{v}_\mathrm{d}}{\p t}\right|_\mathrm{drag}= -
  \rhog\left.\frac{\p \bm{v}_\mathrm{g}}{\p t}\right|_\mathrm{drag}=
  - \frac{\rhog\rhod}{\left(\rhod+\rhog\right)}\frac{\left(\bm{v}_\mathrm{d} -
    \bm{v}_\mathrm{g}\right)}{\tstop}. 
\end{align}
is the dust-gas friction force per unit volume. Note that $\tstop$ 
differs slightly from the stopping time $\tau_\mathrm{s} =
\tstop\rhog/\left(\rhog+\rhod\right)$ defined in previous studies
\citep[e.g.][]{youdin05a,jacquet11}.

We consider the limit of tightly-coupled dust particles with 
$\tstop\OmK\ll 1 $, where $\OmK$ is the Keplerian orbital
frequency. This allows us to simplify the problem by using the
`terminal velocity approximation' so that the dust-gas velocity
difference is 
\begin{align}
  \bm{v}_\mathrm{d} - \bm{v}_\mathrm{g} = \frac{\nabla
    P}{\rhog}\tstop, 
\end{align}
\citep{youdin05a,jacquet11}.

%In the limit of strong dust-gas drag (stopping time $\tstop\to 0$) 
In this limit of strong dust-gas drag, \cite{laibe14} give a single
fluid description of the mixture (their Eq. 84 --- 85 and Eq. 87):

\begin{align}
  &\frac{D\rho}{Dt} = -\rho\nabla\cdot\bm{v}, \label{masseq}\\ 
  &\frac{D\bm{v}}{Dt} = - \nabla\Phi - \frac{1}{\rho}\nabla  P, \label{momeq}\\ 
  &\frac{D\tepsilon}{Dt} = -\frac{1}{\rho} \nabla \cdot \left(\tepsilon 
  \tstop \nabla P \right),\label{dusteq}  
\end{align}
where $D/Dt \equiv \p_t + \bm{v}\cdot\nabla$ is the Lagragian
derivative \emph{following the mixture} with center-of-mass velocity  
\begin{align}
  \bm{v} \equiv \frac{\rhog\bm{v}_\mathrm{g} + 
    \rhod\bm{v}_\mathrm{d}}{\rho}, 
\end{align}
and 
\begin{align}
  \rho \equiv \rhog + \rhod
\end{align}
is the total mixture density.

In Eq. \ref{dusteq}, 
\begin{align}
  \tepsilon \equiv \frac{\rhod}{\rho}  = \frac{\epsilon}{1+\epsilon} 
\end{align}
is the dust-fraction and $\epsilon=\rhod/\rhog$ is the usual
dust-to-gas ratio. The gas pressure $P$ is given by the chosen 
equation of state described below. Note that Eq. \ref{dusteq} is
equivalent to Eq. 46 of \cite{jacquet11}.

For the disk problem the external potential $\Phi$ is  
that for a star of mass $M_*$, 
\begin{align}\label{thin_disk_potential}
  \Phi(r,z) =-\frac{GM_*}{\sqrt{r^2 + z^2}}\simeq
  -\frac{GM_*}{r}\left(1 - \frac{z^2}{2r^2}\right), 
\end{align}
where $G$ is the gravitational constant, and $(r,\phi, z)$ are
cylindrical co-ordinates centered on the star. 
The second equality is the 
thin-disk approximation for the disk potential, appropriate for
$|z|\ll r$. We shall adopt this approximate potential in order to
obtain analytic expressions for the disk equilibria.

\subsection{Locally isothermal equation of state}\label{loc_iso_eos}
To complete the system of equations we need to specify a gas equation
of state. We start with the 
energy equation for an ideal gas of adiabatic index $\gamma$: 
\begin{align}
  \frac{\p P}{\p t} + \bm{v}_\mathrm{g}\cdot\nabla P = - \gamma P
  \nabla\cdot \bm{v}_\mathrm{g}  + \mathcal{H} - \Lambda 
%\frac{\rho \mathcal{R}}{\mu}\frac{T
%  - T_\mathrm{ref}}{t_\mathrm{cool}}, \label{realenergy}
\end{align}
where $\mathcal{H} = 
(\gamma-1)\tepsilon|\rhog\left(\bm{v}_\mathrm{d}-\bm{v}_\mathrm{g}\right)|^2$
accounts for heating due to dust-gas drag \citep{laibe14}; and 
\begin{align}
  \Lambda = \frac{\rho \mathcal{R}}{\mu}\frac{T
  - T_\mathrm{ref}}{t_\mathrm{cool}}, \label{realenergy}
\end{align}
represents radiative cooling on a characteristic timescale
$\tcool$ that relaxes the gas back to a prescribed temperature profile  
$T_\mathrm{ref}$, where $\mathcal{R},\, \mu$ are  the gas constant and
mean molecular weight of the gas, respectively.

We now simplify the gas energy equation by considering short cooling 
times $\tcool\to 0$, appropriate for the outer parts of an irradiated
protoplanetary disk \citep{chiang97,lin15}. Then the disk temperature
$T = T_\mathrm{ref}$ at all times, and so we may
adopt a locally isothermal equation of state 
\begin{align}\label{eos}
  P = c_s^2(r,z)\rhog = c_s^{2}(1 - \tepsilon)\rho,   
\end{align}
where $c_s(r,z)= \sqrt{\mathcal{R}T_\mathrm{ref}/\mu}$ is a prescribed
sound-speed profile fixed in time.

In practice we consider vertically 
isothermal disks with \begin{align}\label{power_temp}
  c_s^2(r) \propto r^{q},
\end{align}
where $q$ is the power-law index for the disk temperature. For $q=0$
the disk is strictly isothermal.

@@

\subsection{Effective energy equation}
Although we have eliminated the need of an evolutionary energy
equation (Eq. \ref{realenergy}) by choosing a locally isothermal
equation of state (Eq. \ref{eos}), we show that the mixture 
nevertheless obeys an effective evolutionary energy equation. This is
because advection of the dust-fraction, described by
Eq. \ref{dusteq}, can be transformed into an energy-like 
equation. We 
%We show that for a prescribed temperature distribution, the mixture
%obeys an evolutionary energy equation, due to the advection of the
%dust-fraction. 
eliminate $\tepsilon$ in favor of $P$, 
\begin{align*}
  \tepsilon = 1 - \frac{P}{c_s^2(r,z)\rho}, 
\end{align*}
then Eq. \ref{dusteq} becomes
\begin{align}
%\frac{DP}{Dt} 
\frac{\p P}{\p t} + \bm{v}\cdot\nabla P  
&= - P \nabla\cdot\bm{v} + P\bm{v}\cdot\nabla\ln{c_s^2}
                + \mathcal{C},  \label{eff_energy} \\
\mathcal{C}&\equiv c_s^2 \nabla\cdot\left[\tstop\left(1 -
  \frac{P}{c_s^2\rho}\right)\nabla 
  P\right].
\end{align}
Comparing with Eq. \ref{realenergy}, we see that 
Eq. \ref{eff_energy} is the energy equation for an ideal gas of
adiabatic index $\gamma=1$, but now with the imposed temperature
gradient $\nabla c_s^2$ and dust-gas drag $\mathcal{C}$ playing the
role of radiative cooling ($\Lambda$ in 
Eq. \ref{realenergy}).

%In this form, the function $\mathcal{C}$ can be interpreted
%as a cooling term.    
If we denote 
\begin{align}
  \bm{F} \equiv  - \frac{\nabla P}{\rho}
\end{align}
for the pressure forces, then
\begin{align*}
  \mathcal{C} = - c_s^2\nabla \left( \tepsilon \tstop \rho \bm{F}
  \right), 
\end{align*}
which is in the same form as cooling by radiative diffusion. In protoplanetary
disks the heat flux, proportional to $\bm{F}$, is directly radially
outwards and vertically upwards. This is simply a reflection of
particle drift (inwards and downwards). Particle flux into a region contributes
to `cooling' of that region because the effective temperature is
lowered.
